---
title: "Ultramarathon gender trends"
output: md_document
#output: html_notebook
---

```{r, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)

options(dplyr.summarise.inform = FALSE)

source(here::here("scr", "00_main.R"))
```

Anyone who's been at the start line of an trail/ultra race would be well aware that the participants are overwhelmingly white and male. This is something that has been identified more formally in a survey conducted by the University of Central Lancashire (https://www.trailultraproject.com/survey) which aimed to look at the culture and habits of trail & ultra runners in the UK and US. Over 95% of the 1,000 respondents described their ethnicity as 'white', while 40% identified as female. The percentage of female respondants to the survey doesn't directly correspond to the gender differences in the results of ultramarathons as obtained from the [D.U.V](http://statistik.d-u-v.org/), consolidated on Kaggle in the [dataset of ultra marathon running](https://www.kaggle.com/datasets/aiaiaidavid/the-big-dataset-of-ultra-marathon-running); with the differences likely due to the US and UK generally having a far better gender balance to other nations. One other thing to note is that the dataset only contains finishers, and any disproportional DNF's between male and female would impact the overall participation numbers.

On the positive side, the survey shows Courtney Dauwalter and Jasmin Paris as the two most inspirational athletes for the US and UK respectively, 

### Current trends

Firstly, as mentioned, the trend does indicate an increase in the female participation rate in ultramarathon races, but closer to 23% globally in 2022 as opposed to 40% of the survey respondents. The chart shows a few different trend changes, with a gradual increase of participation rate of 0.2% per year until 1986, accelerating to 0.8% until 1998 with current trends seeing an approximate 0.3% per year increase in the female/male ratio of participants.

The bad news is that it'll take another 90 years to achieve a 50/50 female/male distribution globally with current trends.

```{r}
print(genderperc.p)
```

However, some countries are far ahead of the curve for gender distribution than others. The USA 


